---
title: "Common Statistical Tests in R"
author: "Martin Chan"
date: "October 7, 2022"
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
# prettyjekyll::FormatPost("_knitr/association_of_variables_20221007.Rmd")
```

# Introduction

This post will focus on common statistical tests in R that you can perform to understand and validate the relationship between two variables. 

Very statistics 101, you may be thinking. So why? This post is edited from my own notes from learning statistics and R. The primary goal of the post is to be practical enough for myself to refer to from time to time, as I do for all my other tutorials/posts in this blog. The secondary goal is that hopefully, as a side effect, that this post will benefit others who are learning statistics and R too. 

To illustrate the R code, I will also be using a sample dataset `pq_data` from the package [**vivainsights**](https://microsoft.github.io/vivainsights/), which is a cross-sectional time-series dataset measuring the collaboration behaviour of simulated employees in an organization. Each row represents an employee on a certain week, with columns measuring behaviours such as total weekly time spent in email, meetings, chats, and so on. 

# Loading the dataset

The package **vivainsights** is available on CRAN, so you can install this with `install.packages("vivainsights")`.

You can load the dataset in R by calling `pq_data` after loading the **vivainsights** package. Here is a preview of the first ten columns of the dataset using `dplyr::glimpse()`:
```{r message=FALSE, warning=FALSE}
library(vivainsights)

glimpse(pq_data[, 1:10])
```

# Understanding the relation between two variables

One of the most basic tasks in statistics and data science is to understand the relation between two variables. Sometimes the motivation is understand whether the relationship is causal, but this is not always the case (for instance, one may simply wish to test for multicollinearity when selecting predictors for a model).

In our dataset, we have two metrics of interest:
  -  `Multitasking_hours` measures the total number of hours the person spent sending emails or instant messages during a meeting or a Teams call.
  - `After_hours_collaboration_hours` measures the number of hours a person has spent in collaboration (meetings, emails, IMs, and calls) outside of working hours.[^1]

[^1]: See <https://learn.microsoft.com/en-us/viva/insights/use/metric-definitions> for definitions.

Imagine then we have two questions to address: 

1. We suspect that multitasking hours is correlated with after-hours working, as the former could be a symptom of excessive workload or sub-optimal time management which can explain after-hours working. What can we do to understand the relationship between the two? 

1. Do managers multi-task more than individual contributors (IC)? 

To answer these questions, there are a couple of common methods at our disposal. Here is a (non-exhaustive) list:

  - Comparison tests
  - Correlation tests
  - Regression tests
  - Effect Size
  - Statistical power
  - Sample variability

It is worth noting that the first question postulates a relation between two **continuous** variables (multitasking hours, afterhours collaboration), whereas the second question a relation between a **categorical** variable (manager/IC) and a **continuous** variable (multitasking hours). The types of the variables in question help determine which tests are appropriate.

## Comparison tests

Two of the most common comparison tests would be the **t-test** and Analysis of Variance (ANOVA). 

A t-test can be paired or unpaired, where the former is used for comparing the means of two groups in the same population, and the latter for independent samples from two populations or groups. An unpaired (two-sample) t-test is therefore appropriate for the scenario in question two, as managers and ICs are two different populations. 

Since we are interested in the difference between managers and ICs, we will first need to create a factor variable from the data that has only two levels. In the below code, we create a new data frame called `pq_data_grouped` containing a new variable called `ManagerIndicator`. This new variable reclassifies any value that equals either "Manager" or "Manager+" to "Manager", and has only two levels: "Individual Contributor" and "Manager":


```{r message=FALSE, warning=FALSE}
pq_data_grouped <-
  pq_data %>%
  mutate(
    ManagerIndicator =
      ifelse(
        test = SupervisorIndicator %in% c("Manager", "Manager+"),
        yes = "Manager",
        no = SupervisorIndicator
      ) %>%
      factor(levels =
               c("Individual Contributor",
                 "Manager"))
  )
```

Recall also that our dataset `pq_data` is a cross-sectional time-series dataset, which means that for every individual identified by `PersonId`, there will be multiple rows corresponding to different dates. To simplify the dataset so that we are looking at person averages, we can group the dataset by `PersonId` and calculate the mean of `Multitasking_hours`. After this manipulation, `Multitasking_hours` would represent the mean multitasking hours _per person_, as opposed to _per person per week_. Let us do this by extending the pipe:


```{r message=FALSE, warning=FALSE}
pq_data_grouped <-
  pq_data %>%
  mutate(
    ManagerIndicator =
      ifelse(
        test = SupervisorIndicator %in% c("Manager", "Manager+"),
        yes = "Manager",
        no = SupervisorIndicator
      ) %>%
      factor(levels =
               c("Individual Contributor",
                 "Manager"))
  ) %>%
  group_by(PersonId, ManagerIndicator) %>%
  summarise(Multitasking_hours = mean(Multitasking_hours), .groups = "drop")
  
glimpse(pq_data_grouped)
```

Once the data is prepared, the t-test can be run with the base `t.test()` function: 

```{r message=FALSE, warning=FALSE}
t.test(
  Multitasking_hours ~ ManagerIndicator,
  data = pq_data_grouped,
  paired = FALSE
)
```

From the above, 

- `t` represents the t-statistic.
- `df` represents the degree of freedom. 
- `p-value` is the p-value. The value here shows to be significant, as it is smaller than 0.05. 
- the test allows us to reject the null hypothesis that the means of multitasking hours between managers and ICs are the same. 

Note that the t-test used here is the Welch's t-test, which is an adaptation of the Student's t-test in that compares the variance of the two groups, i.e. handling heteroscedasticity.

See <https://www.scribbr.com/statistics/statistical-tests/>.

### Testing for normality

But hang on! One of the assumptions behind the t-test is that the data is normally distributed. To test this, we can use the **Shapiro-Wilk test of normality**, with the function `shapiro.test()`:

```{r message=FALSE, warning=FALSE}
pq_data_grouped %>%
  group_by(ManagerIndicator) %>%
  summarise(
    p = shapiro.test(Multitasking_hours)$p.value,
    statistic = shapiro.test(Multitasking_hours)$statistic
  )
```

As both p-values show up as less than 0.05, the test implies that we should reject the null hypothesis that the data are normally distributed. To confirm, you can also perform a visual check for normality using a histogram or a Q-Q plot. 

```{r message=FALSE, warning=FALSE}
# Multitasking hours - IC
mth_ic <-
  pq_data_grouped %>%
  filter(ManagerIndicator == "Individual Contributor") %>%
  pull(Multitasking_hours) 

qqnorm(mth_ic, pch = 1, frame = FALSE)
qqline(mth_ic, col = "steelblue", lwd = 2)

# Multitasking hours - Manager
mth_man <-
  pq_data_grouped %>%
  filter(ManagerIndicator == "Manager") %>%
  pull(Multitasking_hours) 

qqnorm(mth_man, pch = 1, frame = FALSE)
qqline(mth_man, col = "steelblue", lwd = 2)
```

This makes the unpaired t-test less appropriate or conclusive, and we can consider other alternatives, such as the non-parametric two-samples Wilcoxon Rank test. 

### Testing for variance homogeneity

As mentioned above, the difference between the classic Student's t-test and the Welch's t-test is that the former assumes equality between the two variances. Although using the Welch's t-test means that this is not something we have to worry about, here is an example on how to test for variance homogeneity with `var.test()` for comprehensiveness: 

```{r message=FALSE, warning=FALSE}
# F test to compare two variances
var.test(
  Multitasking_hours ~ ManagerIndicator,
  data = pq_data_grouped
  )
```

Note that the code above follows a similar format as `t.test()`. Since the p-value is less than 0.05, we can reject the null hypothesis that variances between the manager and IC dataset are equal. 

## Correlations

Correlation tests are used in statistics to measure how strong a relationship is between two variables, without hypothesising any causal effect between the variables. There are several types of correlation coefficients (e.g. Pearson's _r_, Kendall's _tau_, Spearman's _rho_), but the most commonly used is the Pearson's correlation coefficient.

Correlation tests are a form of **non-parametric test**, which don't make as many assumptions about the data and are useful when one or more of the common statistical assumptions are violated. However, the inferences they make aren't as strong as with parametric tests.

Correlation is a way to test if two variables have any kind of relationship, whereas p-value tells us if the result of an experiment is statistically significant. 


The statistical significance of a correlation can be evaluated with the t-statistic. This can be yielded with `cor.test()` in R:

```r
cor.test()
```

This output provides the correlation coefficient, the t-statistic, df, p-value, and the 95% confidence interval for the correlation coefficient. The two variables you supply to the function should both be continuous (most likely type `numeric`, `integer`, or `double` in R). 

Note that the t statistic for the correlation depends on the magnitude of the correlation coefficient (r) and the **sample size**. With a large sample, even weak correlations can become statistically significant.

### Relationship with simple linear regression

One might also be led to believe that the correlation coefficient is similar to the slope of a simple linear regression. For one, the test for correlation will more or less lead to a similar conclusion as the test for slope. The sign of the slope (+/-ve) will be the same for correlation, and both values should indicate the direction of the relationship. 

However, those two statistics are different: the correlation coefficient only tells you how closely your data fit on a line, so two datasets with the same correlation coefficient can have very different slopes. In other words, the value of the correlation indicates the _strength_ of the linear relationship. The value of the slope does not. Moreover, the slope interpretation tells you the change in the response for a one-unit increase in the predictor. Correlation does not have this kind of interpretation.


### Pitfalls? 

Outliers and non-linear relationships. 

A simple way to evaluate whether a relationship is reasonably linear is to examine a scatter plot.

P-value evaluates how well your data rejects the **null hypothesis**, which states that there is no relationship between two compared groups.

p-value is the probability of obtaining results as extreme or more extreme, given the null hypothesis is true.

## Effect size



## Statistical power

**Statistical power** is the probability of identifying an interaction effect on a dependent variable with the specified sample characteristics. The most common use of power calculations is to estimate how big a sample you will need.

## Sample variance

Examining in-sample variation and between-sample variation are both helpful when comparing means across two populations. 

The f-statistic is in fact calculated by the between-group variance divided by the within-group variance. 

This is where **ANOVA** (Analysis of Variance) comes into play here, where ANOVA is a statistical test used to analyze the difference between the means of more than two groups. The simple version is the one-way ANOVA, but you would also have come across two-way ANOVA which is used to estimate how the mean of a quantitative variable changes according to the levels of two categorical variables (e.g. rain/no-rain and weekend/weekday to predict ice cream sales).

There are three assumptions in ANOVA:
  - The responses for each factor level have a normal population distribution.
  - These distributions have the same variance.
  - The data are independent.

The p-value is found using the f-statistic and the f-distribution.   

## References

- [Correlation and Regression](https://sphweb.bumc.bu.edu/otlt/MPH-Modules/PH717-QuantCore/PH717-Module9-Correlation-Regression/index.html)
- [PennState STAT500](https://online.stat.psu.edu/stat500/lesson/9/9.4/9.4.2)
- [Guide on when to use which statistical tests and when](https://www.scribbr.com/statistics/statistical-tests/)
- [Unpaired t-tests in R](http://www.sthda.com/english/wiki/unpaired-two-samples-t-test-in-r/)