---
title: "Vignette: Simulating a minimal SPSS dataset from R"
author: "Martin Chan"
date: "April 22, 2020"
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
# prettyjekyll::FormatPost("_knitr/First_Post_13-04-19.Rmd")
```

## Background

Simulating data is one of the most useful skills to have in R. For one, it is helpful when you're debugging code, and you have to create a **reprex** (reproducible example) so you can communicate your problem to others for help.[^1] Whether you're a researcher or a business analyst, the data associated with your code is likely to be either confidential so you cannot share it on [Stack Overflow](https://stackoverflow.com/), or way too large or complex for you to upload anyway. Creating an example dataset from a few lines of code which you can safely share is an effective way to get around this problem. 

[^1]: Check out [this RStudio Community thread](https://community.rstudio.com/t/faq-whats-a-reproducible-example-reprex-and-how-do-i-do-one/5219) to find out more about **reprex** (Allegedly, the portmanteau _reprex_ is coined by [Romain Francois](https://twitter.com/romain_francois/status/530011023743655936))

Data simulation is slightly less straightforward with **survey datasets**, which are characterised by (1) variables having labels attached (for both the variable itself and the codes), and (2) a large proportion of ordinal / categorical variables. For instance, a Net Promoter Score (NPS) question/variable is usually accompanied with the variable label "On a scale of 0-10, how likely are you to recommend X to a friend or family?", and is itself an instance of an ordinal variable. If you are trying to reproduce an example

As a bonus, it would also be great to simulate a labelled dataset where there is some 'meaningful' relationship between variables, rather than them being completely random. 

In this post, I will simulate a _minimal_ labelled survey dataset that can be exported to a SPSS (.SAV) file, with full variable and value labels. 

## Getting started

To run this example, we’ll need to load {tidyverse}, {surveytoolbox}, and {haven}. In particular, I’m using {tidyverse} for its data manipulation functions, {surveytoolbox} for functions to set up variable/value labels, and finally {haven} to export the data as a .SAV file.

Note that {surveytoolbox} is currently not available on CRAN yet, but you can install this by running `devtools::install_github("martinctc/surveytoolbox")`.

In addition to loading the packages, let us also set the seed so that the simulated numbers can be reproducible:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(surveytoolbox) # Install with devtools::install_github("martinctc/surveytoolbox")
library(haven)

set.seed(100) # Enable reproducibility
```

## Create individual vectors
To make it easier to control for the individual variables of the labelled dataset, I will take the approach of first setting up individual variables as labelled vectors, and then binding them together at the end into a data frame.   

To start with, I will create `v_id` as an ID variable running from 1 to 1000, which can simply be generated with the `seq()` function. I can then use `set_varl()` from {surveytoolbox} to set a variable label for the `v_id` vector. 


```{r message=FALSE, warning=FALSE}
#### Create individual vectors ####
## Record Identifier
v_id <-
  seq(1, 1000) %>%
  set_varl("Record Identifier")

## Gender
v_gender <-
  sample(1:3, 1000, replace = TRUE, prob = c(.48, .48, .4)) %>%
  set_vall(value_labels = c("Male" = 1,
                            "Female" = 2,
                            "Other" = 3)) %>%
  set_varl("Q1. Gender")
```

The next step is a bit more tricky. I wanted to create a 5-point scale KPI variable (which could represent _customer satisfaction_ or _likelihood to recommend_. One way to do this is to simply run `sample()` again:

```{r message=FALSE, warning=FALSE}
## KPI - #1 simple sampling
v_kpi <-
  sample(x = 1:5,
         size = 1000,
         replace = TRUE) %>%
  set_vall(value_labels = c("Extremely dissatisfied" = 1,
                            "Somewhat dissatisfied" = 2,
                            "Neither" = 3,
                            "Satisfied" = 4,
                            "Extremely satisfied" = 5)) %>%
  set_varl("Q2. KPI")
```

Whilst the above approach is straightforward, the downside is that the numbers are likely to look completely random if we try to actually analyse the results (which is what `sample()` is supposed to do).

I wanted to simulate numbers that are more realistic, i.e. data which will form a discernible pattern when grouping and summarising by gender. What I'll therefore do is to iterate through each number in `v_gender`, and sample numbers based on the gender of the 'respondent'. 

The `prob` argument within `sample()` are completely made up, but are designed to generate results where a bigger KPI value is more likely if `v_gender == 1`, followed by `v_gender == 3`, then `v_gender == 2`.

```{r message=FALSE, warning=FALSE}
## KPI - #2 gender-dependent sampling
v_kpi <-
  v_gender %>%
  map_dbl(function(x){
    if(x == 1){
      sample(1:5,
             size = 1,
             prob = c(2, 3, 3, 5, 5))
    } else if(x == 2){
      sample(1:5,
             size = 1,
             prob = c(2, 4, 5, 4, 3))

    } else {
      sample(1:5,
             size = 1,
             prob = c(2, 3, 3, 4, 3))
    }
  }) %>%
  set_vall(value_labels = c("Extremely dissatisfied" = 1,
                            "Somewhat dissatisfied" = 2,
                            "Neither" = 3,
                            "Satisfied" = 4,
                            "Extremely satisfied" = 5)) %>%
  set_varl("Q2. KPI")
```

For completeness...

```{r message=FALSE, warning=FALSE}
## Outcome KPI - NPS
v_nps <-
  v_gender %>%
  map_dbl(function(x){
    if(x == 1){
      sample(0:10,
             size = 1,
             prob = c(1, 2, 2, 5, 4, 3, 5, 8, 9, 8, 9))

    } else if(x == 2){

      sample(0:10,
             size = 1,
             prob = c(1, 3, 4, 3, 2, 2, 2, 7, 6, 5, 6))

    } else {
      sample(0:10,
             size = 1,
             prob = c(1, 2, 2, 3, 2, 4, 3, 7, 8, 9, 8))
    }
  }) %>%
  as_nps() %>%
  set_varl("Q3. NPS")
```

## Combine vectors

```{r message=FALSE, warning=FALSE}
#### Combine individual vectors ####
combined_df <-
  tibble(id = v_id,
         gender = v_gender,
         kpi = v_kpi,
         nps = v_nps)

```

Looking at the result

```{r message=FALSE, warning=FALSE}
combined_df %>% varl_tb()
```

```{r message=FALSE, warning=FALSE}
combined_df %>%
  select(-id) %>%
  data_dict()
```

```{r eval=FALSE}
combined_df %>%
  sjPlot::view_df()
```

```{r eval=FALSE}
combined_df %>% haven::write_sav("Simulated Dataset.sav")
```

