---
title: "Vignette: Dealing with SPSS labels in R"
author: "Martin Chan"
date: "May 18, 2019"
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(prettyjekyll)
# prettyjekyll::FormatPost("_knitr/First_Post_13-04-19.Rmd")
```

## Background

As someone who works with survey data most of the time, I'm always on the look out for new and better ways to improve my survey analysis workflow in R. Funnily enough, I didn't think R was at all intuitive or easy to work with survey data when I was first starting out to use R. Rather painful, if I'm completely honest!

One of the big reasons was because most survey data cannot really be analysed independently of the **variable** (e.g. *Q1. What is your gender?*) and **value labels** (e.g. 1 = **Male**, 2 = **Female**, 3 = **Other**, ...), and the base **data frame** in R does not easily lend itself to work easily with these labels. A lot of merging, sorting, recoding etc. therefore is then necessary in order to turn the analysis into neat output contingency tables that you typically get via other specialised survey analysis software, like SPSS or Q. Another big reason was of course due to my own ignorance at the time of all the different methods and packages out there which would otherwise make a lot of this easier. 

This post provides a tour of the various functions (from different packages) that I wish I had known at the time. Despite the title, it's not just about **SPSS**: there are plenty of other formats (e.g. SAS files) out there which carry variable and value labels, but I think this title is justified because:

1. Most people starting out on survey data analysis will tend to first come across SPSS files (.sav) 
2. It's still one of the most popular data formats for survey data
3. It's a SPSS file that I will use as a demo in this post (perhaps not a compelling reason, but still a reason)

## Let's start!

Let us first load in all the packages that we'll use in this post. For clarity, I will still make the package-source of the functions explicit (e.g. `labelled::val_label()`) so it's easy to see where each function comes from. One of these packages  **surveytoolbox** is my own and available on Github only, and if you're interested you can install this by running `devtools::install_github("martinctc/surveytoolbox")`.

```{r message=FALSE, warning=FALSE, include=TRUE}
library(tidyverse)
library(haven)
library(sjlabelled)
library(labelled) 
library(surveytoolbox) # install with devtools::install_github("martinctc/surveytoolbox")
```

Since all I really needed is just an open-source, simple, and accessible SPSS / .sav dataset with variable and value labels that I could use for examples, I simply went online and found the first thing that matched these criteria. Not the most exciting - it's the **1991 General Social Survey**, which is nationally representative sample of adults in the United States. You can download the SAV file from the ARDA site [here](http://www.thearda.com/Archive/Files/Downloads/GSS1991_DL2.asp).

`haven::read_sav()` is my favourite way of loading in SPSS files. In my experience, it rarely has any problems and is generally fast enough. There are other alternatives such as `sjlabelled::read_spss()` and `foreign::read.spss()`, which I don't think are practically very different - you can pick a favourite and have these available in your backpocket in case one fails.

```{r echo=TRUE, message=FALSE, warning=FALSE}
source_data <- haven::read_sav("../datasets/General Social Survey_1991.SAV")
```

Running `glimpse()` on the first twenty rows show that many variables are of the **labelled double** class - meaning that these variables would have labels associated with the numeric values they hold. The numbers alone us tell us nothing about what they represent, as these are likely to be categorical variables "in reality".
```{r echo=TRUE, message=FALSE, warning=FALSE}
source_data %>% head(20) %>% glimpse()
```

